#Reproducible Research Course Project 1
Amrit Juneja
==============================================================================

##Read-Me
This markdown file has been divided into the 9 questions as laid out in the assignment. For better continuity of the code, some of the questions have been clubbed together and this has been highlighted wherever applicable.

## 1. Loading and Processing the Data
I have already unzipped the file and this section reads the csv, checks the dimensions and transforms the data as per requirement

The first is to set a global option to turn off knitr warnings and loading the ggplot2 library

```{r, echo=TRUE}
knitr::opts_chunk$set(warning=FALSE)
library(ggplot2)

activity <- read.csv("./activity.csv")
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
weekday <- weekdays(activity$date)
activity <- cbind(activity,weekday)
summary(activity)
```

## 2. Histogram and 3. Mean and Median of steps taken per day
There are 3 parts to this section which are solved for by the code that follows:
1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

Since the mean and the median calculations return two large tables the results are not shown.


```{r, echo=TRUE}
activity_total_steps <- with(activity, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
names(activity_total_steps) <- c("date", "steps")
hist(activity_total_steps$steps, main = "Total number of steps taken per day", xlab = "Total steps taken per day", col = "darkblue", ylim = c(0,20), breaks = seq(0,25000, by=2500))
mean(activity_total_steps$steps)
median(activity_total_steps$steps)
```

## 4. Time series plot of steps taken per day
The following set of code performs makes a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo = TRUE}
average_daily_activity <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)
names(average_daily_activity) <- c("interval", "mean")
plot(average_daily_activity$interval, average_daily_activity$mean, type = "l", col="darkblue", lwd = 2, xlab="Interval", ylab="Average number of steps", main="Average number of steps per intervals")
```

## 5. The 5-minute interval which, on average across all the days in the dataset, contains the maximum number of steps
```{r, echo=TRUE}
average_daily_activity[which.max(average_daily_activity$mean), ]$interval
```

## 6. Code to describe and show a strategy for imputing missing data and 7. Histogram of the total number of steps taken each day after missing values are imputed
In the dataset there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data. The following set of code performs these steps:

1. Calculates and reports the total number of missing values in the dataset (i.e. the total number of rows with NAs)
2. Replaces all the missing values with the average daily mean steps for that day
3. Creates a new dataset that is equal to the original dataset but with the missing data filled in
4. Makes a histogram of the total number of steps taken each day and calculates and reports the mean and median total number of steps taken per day. Also checks how these values differ from the estimates from the first part of the assignment and the impact of imputing missing data on the estimates of the total daily number of steps

```{r, echo=TRUE}
sum(is.na(activity$steps))
imputed_steps <- average_daily_activity$mean[match(activity$interval, average_daily_activity$interval)]
activity_imputed <- transform(activity, steps = ifelse(is.na(activity$steps), yes = imputed_steps, no = activity$steps))
total_steps_imputed <- aggregate(steps ~ date, activity_imputed, sum)
names(total_steps_imputed) <- c("date", "daily_steps")
hist(total_steps_imputed$daily_steps, col = "darkblue", xlab = "Total steps per day", ylim = c(0,30), main = "Total number of steps taken each day", breaks = seq(0,25000,by=2500))
mean(total_steps_imputed$daily_steps)
median(total_steps_imputed$daily_steps)
```

Filling the NA makes the distribution more homogeneous. Moreover, this operation could hide interesting patterns such as the inactivity during particular days of the week. Also, new biases patterns evidently appear in the calculation operated on the activity2 dataset.

## 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends and 9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report
There are 2 steps to this problem:
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day
2. Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data

```{r,echo=TRUE}
activity$weekday <- factor(format(activity$date, "%A"))

levels(activity$weekday) <- list(weekday = c("Monday", "Tuesday",
                                              "Wednesday", "Thursday",
                                              "Friday"), weekend =
                                          c("Saturday", "Sunday"))
par(mfrow = c(2, 1))

with(activity[activity$weekday == "weekend",], plot(aggregate(steps ~ interval, FUN = mean), type = "l", main = "Weekends"))

with(activity[activity$weekday == "weekday",], plot(aggregate(steps ~ interval, FUN = mean), type = "l", main = "Weekdays"))
```

